United Monoids

### 2021-10-04 paper/main.tex
Expand abstract following the guidelines

\begin{abstract}
This paper demonstrates how a simple inductive data type can be used as a
language for describing and manipulating various combinatorial graph-like
structures. By changing the rules of interpretation of the language, we will
obtain sets, preorders, edge-labelled graphs, and simplicial sets.

As a theoretical contribution, the paper presents united monoids, an algebraic
structure that turns out to be a common ground for the graph-like structures
discussed in the paper.
\end{abstract}

>>>>

\begin{abstract}
% Below is a shorter version of the abstract, which I prefer. Alas, it does not
% conform to the journal guidelines.
%
% =================================
% This paper demonstrates how a simple inductive data type can be used as a
% language for describing and manipulating various combinatorial graph-like
% structures. By changing the rules of interpretation of the language, we will
% obtain sets, preorders, edge-labelled graphs, and simplicial sets.
%
% As a theoretical contribution, the paper presents united monoids, an algebraic
% structure that turns out to be a common ground for the graph-like structures
% discussed in the paper.
% =================================

% Guidelines specify the following structure:
%
% Context: What is the broad context of the work? What is the importance of the
% general research area?
%
% Inquiry: What problem or question does the paper address? How has this problem
% or question been addressed by others (if at all)?
%
% Approach: What was done that unveiled new knowledge?
%
% Knowledge: What new facts were uncovered? If the research was not results
% oriented, what new capabilities are enabled by the work?
%
% Grounding: What argument, feasibility proof, artifacts, or results and
% evaluation support this work?
%
% Importance: Why does this work matter?

\emph{Context:} Graphs and various graph-like combinatorial structures, such as
preorders and hypergraphs, are ubiquitous in programming. This paper focuses on
representing graphs in a purely functional programming language like Haskell.
There are several existing approaches; one of the most recently developed ones
is the ``algebraic graphs'' approach (2017). It uses an algebraic data type to
represent graphs and has attracted users, including from industry, due to its
emphasis on equational reasoning and making a common class of bugs impossible
by eliminating internal invariants.

\emph{Inquiry:} The previous formulation of algebraic graphs did not support
edge labels, which was a serious practical limitation. This paper redesigns the
main algebraic data type and removes this limitation.

\emph{Approach:} We follow a fairly standard approach of parameterising a data
structure with a semiring of edge labels. The newly developed data structure is
both more general and also simpler: the two ways of composing graphs used in the
original formulation can now be obtained from a single operation by fixing the
semiring parameter to zero and one, respectively.

\emph{Knowledge:} By analysing the resulting data structure and formulating laws
for interpreting expressions of the corresponding type, we discover an unusual
algebraic structure, which we call ``united monoids'', that is, a pair of
monoids whose unit elements coincide. We believe that it is worth studying
united monoids in their full generality, going beyond graphs which prompted
their discovery. To that end, we characterise united monoids with a minimal set
of axioms, prove a few basic theorems, and discuss several notable examples.

\emph{Grounding}: We validate the presented approach by implementing it in the
open-source \textsf{algebraic-graphs} library. Our theoretical contributions
are supported by proofs that are included in the paper itself and are also
available in a machine-checked artifact.

\emph{Importance:} By extending algebraic graphs with support for edge labels,
we make them suitable for a much larger class of possible applications. By
studying united monoids, we provide a theoretical foundation for further
research in this area.
\end{abstract}
